Dancing on Water: The Choreography of Sulfur Dioxide Adsorption to Aqueous Surfaces

### 2011-10-19 ReviewResponses.tex

The authors greatly appreciate the detail of the reviews provided, and have made significant changes to the manuscript to incorporate the suggestions, and to clarify the various points raised.

Review #3

1. The suggested literature (Garrett et al. and Wilson) cover many aspects of molecular dynamics simulations and computational techniques used for analysis of gaseous uptake into aqueous surfaces. The work of Garrett, Schenter, and Morita address the landscape of current techniques and water models used, but do not directly address sulfur dioxide uptake in their included analyses. The referenced publications in that work have already been included in our manuscript (see Jayne et al, 1990). Wilson and Pohorille (J. Phys. Chem. B 1997, 101, 3130) look at the adsorption of alcohols to the liquid-vapor interface, and present a refined analysis of the solvation process, but do not address sulfur dioxide uptake or orientation. We believe that although dozens, if not hundreds, of very fine publications have been dedicated to uptake of various species at the air-water interface, referencing extraneous works that do not specifically address sulfur dioxide behavior would detract from our present manuscript. We did not aim for this manuscript to be an exhaustive review of the field of gaseous uptake and solvation processes via molecular dynamics simulations. We have thus limited the scope of our bibliography to include only the most pertinent publication relating to sulfur dioxide behavior.

We believe that our introduction to the manuscript specifically addresses what is and is not known, and how the present work adds to our knowledge of sulfur dioxide behavior. We emphasize here again that our study was not set to be a review of the field of gaseous uptake, but a narrowly defined study of sulfur dioxide orientational behavior on an aqueous surface. Thus we did not include a discussion relating to other gaseous molecules or related systems.

2. The suggestion to use bivariate distributions is appreciated. We have reworked our data and presented the relevant depth-profiles as bivariate plots P($\theta$, $\phi$) showing the specific orientations of water and sulfur dioxide at various depths (slices) throughout the interfacial region.

3. The "Computational Approach" section does not contain any text regarding results or analysis. The section is reserved to define the computational approach, and to define the terminology used regarding depth and orientational analysis.

4. The term "absorption" is used only once in the computational methods section. Regarding the abstract, we have submitted a corrected version.

5. To address the used of the POL3 model with the SO2 model developed by Baer et al, we have submitted a supplemental analysis to be included with the manuscript. We have calculated the radial distribution function of the water-SO2 system to compare the resultant structure and geometries of the interactions between the two types of molecules. Furthermore, we have calculated the SO2-H2O interaction as a function of the S-O separation. Both of these are directly comparable to the results of Baer et al. Our calculations show a nearly identical result to Baer's in both the interaction energy and liquid structure using the POL3 water model. Thus we feel this justifies the use of the more computationally efficient POL3 model.

6. The Henry's law constant has been added to the text.

7. The pressure in the gas region was set initially to 1 atm. The system was then equilibrated until all the gaseous SO2 adsorbed to the water surface, at which point another 1 atm of SO2 was introduced. This procedure was repeated until the gas phase SO2 no longer adsorbed to the surface. This required a total of 50 SO2 molecules introduced in the gas phase. Thus a constant pressure of 1 atm was used throughout equilibration and data collection simulations.

8. We have addressed the technical details in the text.

9. Sections for both the equilibrium MD and the steered MD were specifically included to resolve any doubt regarding the simulations of the two different systems. The text may appear to be redundant, but we are addressing two separate systems that are simulated slightly differently.

10. The terminology has been chosen as the most appropriate descriptor for the system to distinguish the equilibrated system from the steered MD system. System descriptors have been used consistently throughout the manuscript, and we feel that they are appropriate to the current work.

11a. The system was simulated using the NSCM flag in Amber, and the drift still appears. This is not a drift of the entire system, but a slight swelling or movement of the slab over time. The problem is not remedied by the suggested solution.

11b. The same calculation was applied to both surfaces. The surface location was determined as suggested by the reviewer: independently for both surfaces.

12. Text regarding the Gibb's dividing surface has been updated to correct the language used. The manuscript now referes to the width of the water density profile.

13. Hydrogen bond lengths were taken from the radial distribution functions of interfacial waters. The manuscript has been updated by removing the note regarding the "typical bond length".

14c. The profiles were indeed averaged over both interfaces of each simulated slab. It has been noted in the manuscript.

14d. The water penetration into the SO2 layer is discussed at length later in the manuscript. A few sentences regarding this behavior have been added to the section on the density profiles.

15. We have removed this section from the manuscript as we are attempting to limit the scope of our work to the study of SO2 behavior, and not as a review of water models used in various capacities. The results of the current work for the POL3 model used agree well with our previous MD studies employing POL3. We have added in the supplemental section our justification for the use of this model.

### 2011-10-19 Introduction.tex

An important gas for developing a picture of gaseous adsorption and entry into a water surface is sulfur dioxide.\cite{Lattanzi2010,Shah2011,Tzivian2011,Johns2011,Faloona2009,Jurkat2010,Wu2011,Jayne1990,Yang2002} \suldiox~enters the environment as an important industrial product, and also naturally through terrestrial processes. Atmospheric dust particles and gases have been implicated in the oxidation of \suldiox, and act as reaction surfaces for chemical mechanisms that are still poorly understood.\cite{Baltrusaitis2011,Rubasinghege2010,Li2007,Madsen2008,Boniface2000} \suldiox~acts as a major component of atmospheric pollution, and is a precursor to acid rain formation, and cloud nucleation. Its high solubility in water makes \suldiox~an integral compound in many aqueous atmospheric reactions, as well.  Obtaining a more complete picture of the \suldiox~adsorption process is important for understanding gaseous adsorption of this environmentally important gas on water and aerosol surfaces as well as being a model system for understanding the more general nature of gases at aqueous interfaces.

>>>>

An important gas for developing a picture of gaseous adsorption and entry into a water surface is sulfur dioxide.\cite{Lattanzi2010,Shah2011,Tzivian2011,Johns2011,Faloona2009,Jurkat2010,Wu2011,Jayne1990,Jayne1990a,Yang2002} \suldiox~enters the environment as an important industrial product, and also naturally through terrestrial processes. Atmospheric dust particles and gases have been implicated in the oxidation of \suldiox, and act as reaction surfaces for chemical mechanisms that are still poorly understood.\cite{Baltrusaitis2011,Rubasinghege2010,Li2007,Madsen2008,Boniface2000} \suldiox~acts as a major component of atmospheric pollution, and is a precursor to acid rain formation, and cloud nucleation. Its high solubility in water makes \suldiox~an integral compound in many aqueous atmospheric reactions, as well.  Obtaining a more complete picture of the \suldiox~adsorption process is important for understanding gaseous adsorption of this environmentally important gas on water and aerosol surfaces as well as being a model system for understanding the more general nature of gases at aqueous interfaces.

### 2011-10-19 Methods.tex

Molecular dynamics simulations were performed using the Amber 11 software suite.\cite{Case2010} Polarizable models for the \wat~and \suldiox~molecules were used in the simulations, and have been used previously in studies on interfacial systems because they are known to more accurately reproduce interfacial structure and free energy profiles.\cite{Wick2007,Rivera2006,Dang1998} The \wat~model used is the POL3 water model,\cite{Caldwell1995} and for \suldiox~we used the model of Baer et al. that places a single polarizable center on the sulfur atom.\cite{Baer2010}

@@

Figure \ref{fig:density-flaw} demonstrates the problem of surface location drift during a simulation. Figure \ref{fig:density-flaw}A shows the density profile of water and \suldiox~over the course of one of the 10 ns trajectories used in this work using the original uncorrected coordinates of the system taken from the raw atomic positions of the molecular dynamics data output. The water density profile and location (thin gray line in Figure \ref{fig:density-flaw}A) is produced by averaging the instantaneous density profile at each time step in the simulation over all the time steps. The water profile was then fit to a $\tanh$ function (black line) to extract the position and width parameters of the water surface. The fitted Gibbs dividing surface has a width of 3.77\angs, which is comparable to values reported for similar neat-water systems.\cite{Dang1997,Hore2008} A bulge in the gas-phase ($\leq 52$\angs) side of the water density profile is indicative of the drift of the water slab over the course of the trajectory. Thus the calculated location and width from the $\tanh$ line fit are not accurate over long trajectories for defining a stationary reference point.

	To overcome this problem we define the water surface location by calculating a reference location at each time step by averaging the positions of the waters contained in the topmost monolayer. This provides a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden. The number of waters included in the averaging is determined by taking a few issues into account. First, counting the waters found in the topmost cross-section of the water slab over several time steps indicated between 65-75 waters that established a full monolayer. This was done by a visual inspection of the slab using the VMD MD visualization package.\cite{Humphrey1996} Alternatively, assuming a spherical model of water with a radius of 2.2\angs~(the typical length of a bulk-water hydrogen bond), two layers of hexagonally tight-packed spheres yielded a similar number of surface water molecules. Increasing the number of waters used in calculating the surface location diminishes the effects of the few waters that briefly rise above the surface into the gas phase, stabilizing both the surface position and thickness values. Taking the close-packed model as a maximum number of waters fit into a flat surface, the topmost 70 water molecules were used for calculating instantaneous water surface locations for each simulation step.

>>>>

Molecular dynamics simulations were performed using the Amber 11 software suite.\cite{Case2010} Polarizable models for the \wat~and \suldiox~molecules were used in the simulations, and have been used previously in studies on interfacial systems because they are known to more accurately reproduce interfacial structure and free energy profiles.\cite{Wick2007,Rivera2006,Dang1998} The \wat~model used is the POL3 water model,\cite{Caldwell1995} and for \suldiox~we used the model of Baer et al. that places a single polarizable center on the sulfur atom.\cite{Baer2010} An intermolecular cutoff of 12\angs~was used for long-range electrostatic forces. The simulations were performed in the NVT ensemble using Langevin dynamics for temperature control. Induced dipoles were treated by the polarizable potential functions of the Amber molecular dynamics software.

@@

Figure \ref{fig:density-flaw} demonstrates the problem of surface location drift during a simulation. Figure \ref{fig:density-flaw}A shows the density profile of water and \suldiox~over the course of one of the 10 ns trajectories used in this work using the original uncorrected coordinates of the system taken from the raw atomic positions of the molecular dynamics data output. The water density profile and location (thin gray line in Figure \ref{fig:density-flaw}A) is produced by averaging the instantaneous density profile at each time step in the simulation over all the time steps. The water profile was then fit to a $\tanh$ function (black line) to extract the position and width parameters of the water surface. The fitted surface water density profile has a width of 3.77\angs, which is comparable to values reported for similar neat-water systems.\cite{Dang1997,Hore2008} A bulge in the gas-phase ($\leq 52$\angs) side of the water density profile is indicative of the drift of the water slab over the course of the trajectory. Thus the calculated location and width from the $\tanh$ line fit are not accurate over long trajectories for defining a stationary reference point.

	To overcome this problem we define the water surface location by calculating a reference location at each time step by averaging the positions of the waters contained in the topmost monolayer. This provides a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden. The number of waters included in the averaging is determined by taking a few issues into account. First, counting the waters found in the topmost cross-section of the water slab over several time steps indicated between 65-75 waters that established a full monolayer. This was done by a visual inspection of the slab using the VMD MD visualization package.\cite{Humphrey1996} Alternatively, assuming a spherical model of water with a radius of 2.2\angs, two layers of hexagonally tight-packed spheres yielded a similar number of surface water molecules. Increasing the number of waters used in calculating the surface location diminishes the effects of the few waters that briefly rise above the surface into the gas phase, stabilizing both the surface position and thickness values. Taking the close-packed model as a maximum number of waters fit into a flat surface, the topmost 70 water molecules were used for calculating instantaneous water surface locations for each simulation step.
